Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-11 paper/relatedwork.tex
Pound on program bifurcation really hard in the related work.

Frequency reduction and precomputation are common techniques for both
designing algorithms and performing compiler
optimizations~\cite{JS86-staging}.
%
The idea behind precomputation is to identify computations that can be
performed earlier (e.g., at compile time) if their inputs are
available statically and perform them at that earlier time. Dynamic
algorithms, partial evaluation, and incremental computation are all
examples of precomputation techniques.
%
The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  Dynamic programming, loop hoisting, and pass
separation (presented here) are all examples of frequency reduction
techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (runtime) stages. Given a program and values 
for all static inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52}.
%Kleene did not intend to improve efficiency, however.  Lombardi
%appears to be the first person to use the term in his work on
%incremental computation~\cite{Lombardi67}.
There has since been an enormous body of work on partial evaluation,
and we refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of work until the early 90's~\cite{JGS93}.

@@

\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
is a metastatic staging transformations
used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
Bifurcation is similar to stage splitting in that it factors the input program
into two versions based on the stages of the input.
However, in bifurcation there is no building of data structures to be use in the resumer;
instead, the second stage version of the code also has 
access to the first stage inputs and just recomputes whatever it needs.
That difference aside, the grammar-based binding-time specifications used in bifurcation
were capable of describing data structures with purely-static, purely-dynamic, and mixed stage content,
much like the type system of \lang.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance was difficult to
predict. This led to investigations based on type systems for making
the stage of computations explicit in
programs~\cite{GJ91-lambda,NN92-twolevel} and writing ``metaprograms''
that, when evaluated at a stage, yield a new program to be evaluated
at the next stage.
%J{\o}rring and
%Scherlis describe the idea of meta programming as one of the three possible approaches to
%taking advantage of staging~\cite{JS86-staging}.
Davies~\cite{davies96} presented a
logical construction of binding-time type systems by deriving a type
system via the Curry-Howard correspondence applied to temporal logic.
Davies and Pfenning proposed a new type system for staged computation
based on a particular fragment of modal logic~\cite{DP01-modal}. The
work on MetaML extended type-based techniques to a full-scale language
by developing a statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variables to occur within staged
computations~\cite{NP05-nn}.

@@

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where, to achieve high performance,
programs are typically manually separated into components by frequency
of execution corresponding to support graphics hardware pipeline
stages. The software engineering challenges of modifying multiple
per-stage programs have led to suggestions of writing graphics
programs in an explicitly-staged programming
language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and deferring the
task of pass separation to the compiler. However, all prior efforts,
like that of Knoblock and Ruf\,\cite{knoblock96} have been limited to
simple, imperative languages.

%% Online and offline.  Both approaches have their merits but online
%% partial evaluation can be too slow and if stopped lead to an
%% inefficient residual.  Offline partial evaluation is faster.  Also,
%% full self-application has only been achieved by offline partial
%% evaluation.

%% Weaknesses of partial evaluation include~\cite{GJ05}:
%% 1) speedups linear in the subject programs run time (excuse moi?)

%% 2) good speedups require close knowledge of the program and some
%% binding-time improvements.

%% 3) result of specialization can be hard to predict as it can lead to
%% slowdown, code explosion, and non-termination at specialization time.

%% \paragraph{Meta programming}

%% What happens when we try to write quickselect as a meta program?

%% \ur{Umut: I am starting a paragraph on termination.  we might need to expand
%%   on this.}
%% \paragraph{Termination.}

%% There are two reasons for non-termination: infinite loops and
%% execution of errorneous code.  The splitting algorithm can cause the
%% first.  But can it also cause the second? (If all computations really
%% depend on stage 1 code, then would it be possible to for example see
%% something like 1/0?  i think so but only if there is a meta dependency
%% between the data in two stages?)

%% Is in possible for the residual program to not terminate (unexpectedly)?

%% Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.

%% Based on Glenstrup and Jones (pp. 1154, 1155), a partial evaluator
%% should be complete. That is every computation depending on the static
%% input should be reduced to a value (during specialization).
%% Unfortunately, completeness is difficult to achieve if we also require
%% that the specialization is always terminating.

%% One problem with their description is that it is not clear which
%% termination problem they are talking about termination of the residual
%% or the specializer.

%% Nontermination during specialization: 
%% consider code

%% double x = doubleplus (x,0)

%% doubleplus (u,v) = if u <= 0 then v else doubleplus (u-1,v+2)

%% this can lead to non-termination if v is considered static.

%% Is this a problem for us?

>>>>

Frequency reduction and precomputation are common techniques for both
designing algorithms and performing compiler
optimizations~\cite{JS86-staging}.
%
The idea behind precomputation is to identify computations that can be
performed earlier (e.g., at compile time) if their inputs are
available statically and perform them at that earlier time. Dynamic
algorithms, partial evaluation, and incremental computation are all
examples of precomputation techniques.
%
The idea behind frequency reduction is to identify computations that
are performed multiple times and pull them ahead so that they can be
performed once and used later as needed.  Dynamic programming, loop
hoisting, and splitting (presented here) are all examples of frequency
reduction techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is partial
evaluation, which distinguishes between static (compile-time) and
dynamic (runtime) stages. Given a program and values for all static
inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52},
%Kleene did not intend to improve efficiency, however.  Lombardi
%appears to be the first person to use the term in his work on
%incremental computation~\cite{Lombardi67}.
%There has since been an enormous body of work on partial evaluation,
and we refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of partial evaluation work until the early
90's~\cite{JGS93}.

@@

In an attempt to reduce the complexity of algorithms for
specialization used in self-applicable partial evaluators, researchers
explored the use of staging transformations that, similar to
splitting, partition an input program into two programs: one that
produces static outputs from static inputs, and a second that produces
dynamic outputs from both static and dynamic inputs.  These metastatic
transformations, referred to as \emph{program
  bifurcation}\,\cite{DeNiel91} or as \emph{binding time
  separation}\,\cite{Mogensen89a}), were successfully applied to
functional languages that supported
recursion\,\cite{DeNiel91}. However, unlike splitting, where the goal
is to emit code where first stage results are computed once and then
reused in multiple invocations of second stage execution, program
bifurcation only serves as a preprocessing step for the subsequent
specialization phase of partial evaluation. The second function
emitted by program bifurcation has access to the first stage inputs
and recomputes first stage computations it needs. (As noted by
Knoblock and Ruf\,\cite{knoblock96}, it may be possible to modify the
program bifurcation algorithm to cache and reuse the intermediate
results, but this was never attempted.) The grammar-based binding-time
specifications used in program bifurcation were capable of describing
data structures with purely-static, purely-dynamic, and mixed stage
content, much like the type system of \lang.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance was difficult to
predict. This led to investigations based on type systems for making
the stage of computations explicit in
programs~\cite{GJ91-lambda,NN92-twolevel} and writing ``metaprograms''
that, when evaluated at a stage, yield a new program to be evaluated
at the next stage.  Davies~\cite{davies96} presented a logical
construction of binding-time type systems by deriving a type system
via the Curry-Howard correspondence applied to temporal logic.  Davies
and Pfenning proposed a new type system for staged computation based
on a particular fragment of modal logic~\cite{DP01-modal}. The work on
MetaML extended type-based techniques to a full-scale language by
developing a statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variables to occur within staged
computations~\cite{NP05-nn}.

@@

As noted in \ref{sec:graphics}, splitting algorithms have also been a
topic of interest in programming systems for computer graphics, where,
to achieve high performance, programs are manually separated into
components by frequency of execution corresponding to graphics
hardware pipeline stages. The software engineering challenges of
modifying multiple per-stage programs have led to suggestions of
writing graphics programs in an explicitly-staged programming
language\,\cite{Proudfoot:2001,Foley:2011,He:2014} and deferring the
task of pass separation to the compiler. However, all prior splitting
efforts in computer graphics, like that of Knoblock and
Ruf\,\cite{knoblock96} have been limited to simple, imperative
languages.
